Data analysis to prioritise new cycleways in response to COVID-19

### 2020-06-02 report.Rmd
Add report

---
title: "Methods to prioritise pop-up active transport infrastructure"
# : overview and findings from 10 UK cities"
author: "Robin Lovelace, Malcolm Morgan, Joey Talbot, Martin-Lucas-Smith"
# output: bookdown::github_document2
# output: bookdown::html_document2
# output:
#   bookdown::pdf_document2:
#     keep_tex: true
# bibliography: references.bib
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, out.width = "100%")
```

```{r, echo=FALSE, eval=FALSE}
citr::tidy_bib_file("article.Rmd", "~/uaf/allrefs.bib", file = "ref.bib")
wordcountaddin::word_count("article.Rmd")
knitr::purl("article.Rmd")
file.copy("article.R", "code/article-methods-1.R")
pagedown::chrome_print("article.html")
browseURL("article.pdf")
zf = list.files(pattern = "tex|png", full.names = TRUE)
zfl = list.files("article_files/figure-latex/", full.names = TRUE)
fz = c(zf, zfl)
msgf = paste(fz, collapse = " ")
msgf = gsub(pattern = "\\./", replacement = "", msgf)
msg = paste("tar -cvf article.tar ", msgf)
system(msg)
```

>>>>

---
title: "Data analysis to prioritise new cycleways in response to COVID-19"
author: Dr Robin Lovelace & Dr Joey Talbot, University of Leeds
output: bookdown::pdf_document2
bibliography: report-refs.bib
---

```{r, eval=FALSE, echo=FALSE}
citr::tidy_bib_file("~/uaf/allrefs.bib", rmd_file = "report.Rmd", file = "report-refs.bib")
citr::tidy_bib_file("article.Rmd", "~/uaf/allrefs.bib", file = "ref.bib")
browseURL("report.pdf")
```

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
```

# Introduction

New statutory guidance and funding for active transport infrastructure was announced by the Secretary of State for Transport on May 9^th^, to support the UK's response to the COVID-19 crisis by enabling healthy and safe travel at a time of reduced capacity on public transport networks.^[
https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19
]

The statutory guidance on network management in response to COVID-19 outlines a range of measures that can support a safe and healthy restart, including the creation of new cycleways, reducing speed limits, widening of footways and closing roads to motor traffic via 'modal filters'.
**This report describes the rapid cycleway prioritisation tool project, which provides aims to ensure evidence-based and effective investment in first of these measures: the creation of new cycleways in existing carriageway space reclaimed from motor traffic.**
Light segregation infrastructure such as flexible plastic wands (illustrated in Figure 1) and bollards can deliver a step change in provision for cycling, when implemented in combination with other measures such reduced speed limits on key corridors (as has happened on the A33 in Southampton and the A56 in Manchester) and when placed where they are most needed.
**The purpose of the tool is to help identify such key corridors, ensuring that funds from the Emergency Active Travel Fund change the status quo of city transport systems in favour of active travel.**

<!-- 17:29:01	 From Gill M (Manchester cyclist) : Trafford is planning to reduce speeds on Bridgewater Way element of A56 when the pop up lane goes in there “OneTrafford tweeted that work is continuing to add pop up cycle lanes along Cornbrook Road and Warwick Road in Stretford too, with temporary 30mph speed limit imposed along Bridgewater Way.” Source (https://www.messengernewspapers.co.uk/news/18489349.cyclists-praise-pop-up-cycle-lanes-along-a56-trafford/) -->

<!-- https://twitter.com/doddyuk/status/1267146449227255808 -->

```{r park-lane, fig.cap="New space for cycling created by road space reallocation in Park Lane, London.", out.width="60%"}
# source: Hillingdon and Uxbridge Times
# https://www.hillingdontimes.co.uk/news/national/18450461.one-londons-prestigious-roads-gets-cycle-lane-due-pandemic/
knitr::include_graphics("figures/cycleway-image.jpg")
```

The £250m released by the Emergency Active Travel Fund will be spent in three broad ways:

- £25m will be invested in cycle repair schemes
- £45m will be invested on swift interventions that must be implemented in June
- £180m will be released later in the summer for more permanent measures

Given the importance of acting quickly, and the unique opportunity that the crisis poses to improve physical activity levels in the long term, it is vital that local authorities have the evidence they need, to ensure the funding is spent where it is most needed.
Without a nationally consistent and systematic evidence base, there is a risk that already stretched local authorities rely on ad-hoc and flawed methods to allocate funding.

In this context, the aim of work outlined in this document is to identify roads and road segments where investment in new cycleways should be prioritised using a transparent methodology and nationally consistent datasets, alongside local authority expertise and local knowledge.
Strong and actionable datasets already exist to support strategic cycle network planning, notably the Propensity to Cycle Tool (PCT), a national cycling research project and planning support system [@lovelace_propensity_2017].
The PCT is already used by local authorities and contains key data showing not only current levels of cycling but also scenarios of growth, notably the Government Target scenario of doubling the number of cycling stages made by 2025 [@departmentfortransport_cycling_2017], at area, route and route network levels, for commuting and school travel (as of May 2020) [@goodman_scenarios_2019].
A limitation of the PCT, however, is that it takes little account of existing infrastructure: variables such as road width and even the presence/absence of cycle infrastructure currently are not included in the results.

To address this limitation, the work set out below will use national road network data which includes geographic attributes and cycle infrastructure from OpenStreetMap, plus carriageway width data derived from analysis of Ordnance Survey data.
This road data was developed for the Cycling Infrastructure Prioritisation Tool (CyIPT) project, funded as a prototype project by a DfT Transport-technology research innovation grants (T-TRIG).
The results of CyIPT, while not publicly accessible, are well known among local authority transport planners and are reported to be of great use to practitioners in scheme prioritisation, according to a DfT funded Discovery project on tools to support investment in active transport.^[
Planning Tools for cycle networks and infrastructure prioritisation, unpublished.
]

In response to the urgent need for evidence on how to deal with the transport-related challenges of COVID-19, and building on our expertise developing nationally scalable tools for active transport planning, we have developed methods to prioritise new cycleways, based on levels of cycling potential *and* road attribute data.
The method was parameterised to highlight roads that have sufficient space to allow for new cycleways without affecting the navigable network for motorised modes, as reported in a paper submitted for peer review in the journal Transport Findings, currently available as a preprint [@lovelace_methods_2020].
A key benefit of this method is that it ranks roads based not only on cycling potential, but also based on continuous road length, following international guidance that emphasises the importance of continuity and cohesion in cycle network design [@parkin_designing_2018].

The method has been tested on 10 major cities in England and feedback from stakeholders with local knowledge secured by CyclingUK has been overwhelmingly positive.
People with good working knowledge of Bristol, Cambridge, Leeds, London, Manchester and Newcastle have provided detailed feedback suggesting that the results match intuition of where new cycleways are most needed.
Several specific requests for results in additional cities have been made, including from practitioners working in 
Coventry, Nottingham, Southampton and Norwich, showing high demand for the evidence provided by our methods in a range of locations.
Building on strong networks with local authorities, we will elicit feedback via the DfT's Basecamp forum, in addition to feedback from other practitioners that we will document in a systematic way.
This will allow us to iterate our methods based on local knowledge, ensuring the results are accurate and of maximum use before being published nationwide.

<!-- Coventry (Adam Tranter), Nottingham (Mark Dimond and Mark Pasley), Southampton (Dr Ben Anderson) -->

<!-- Commenting out further work. Note: this could go into future pitches -->
<!-- Further work is possible. -->
<!-- We could extend the analysis to calculate the 'propensity for mode shift', from public transport modes, resulting in *new cycling potential* estimates that directly account for reductions in public transport use. -->
<!-- We could also incorporate additional variables --- such as population density, proximity to key destinations such as hospitals and major employment centres --- to refine the results to account for non-commute trips. -->
<!-- And we could develop methods to support prioritisation of other types of intervention beyond new cycleway provision via road space reallocation, such as 'filtered permeability' interventions to prevent through traffic in residential areas, greatly improving road safety. -->
<!-- This additional work would take substantially longer than the 3/4 weeks discussed and require more resources.  -->

# Outputs

The main output from this work will be a national evidence-base to support investment in pop-up active transport infrastructure, with results down to the road network level, to support national and particularly local prioritisation of schemes and investment.
Specifically, we will provide:

- A national map of roads that are promising candidates for pop-up active transport infrastructure, that takes into account cycling potential, road space, existing infrastructure and network cohesion
- Tabular and map-based outputs prioritising roads for new cycleways authority, plus open data on road sections highlighted by the analysis
- A report setting out methods, national level results, limitations and guidance on effective use of the results for local authorities
<!-- - (Optional) an online interface that enables users to select the local authority of interest and relevant variables including candidate segments, existing infrastructure and road sections prioritised by our methods -->

<!-- Additional outputs and analysis, for example including additional variables and refining the methods to focus on public transport / cycling mode shift potential, options that have been discussed with the DfT, are possible but not within the timescale of the proposed work. -->

# Tasks and timeline

The main tasks, with estimated resources per task, are outlined below.

- Develop method for automating the generation of results at the unitary authority level (2 days)
- National 'build' and roll-out of methods to cover all authorities (2 days)
- Comparison of results with existing infrastructure to ensure prioritisation of schemes that contribute to cohesive networks (2 days)
- Elicit and incorporate feedback from key stakeholders from a range of local authorities (e.g. including those with urban/rural populations, monocentric/polycentric urban forms and mode split profiles) within a limited timeframe to identify any systemic issues with the results (2 days)
- Geographic analysis to ensure unitary authority boundary definitions are up-to-date and match highway authority definitions (1 day)
- Deploy results in interactive web maps (2 days)
- Write up the work, including description of the methods, overview of the key findings nationwide, assessment of limitations, scope for further work and guidance on effective use of the results (2 days)

<!-- - Optional: provision of results in a web application with an interactive user interface (5 days) -->

We can deliver the project within four weeks, with an estimated timeline as follows:

```{r, echo=FALSE, message=FALSE}
library(gitgantt)
gitgantt::gg_timevis(owner = "cyipt", repo = "popupCyclewaysEngland")
```

# Milestones

First cut of results: Friday 29th

First draft report: Friday 5th

Final write-up of results: Friday 12th

<!-- # Further work -->

<!-- Building on the work outlined in the previous section, further work could be done to generate a more sophisticated evidence base for prioritising active transport infrastructure in the medium-to-long term. -->
<!-- It will take longer than the 3/4 week timescale for rapid decision-making to deliver these tasks and they would require more dedicated researcher time, with estimated days required for key tasks outlined below. -->

<!-- - Inclusion of additional variables such as population density, proximity to key destinations such as hospitals and major employment centres to better account for the specific types of commute and non-commute trip that are currently being undertaken (5 - 20 days) -->
<!-- - Extend the analysis to calculate the 'propensity for mode shift' from public transport modes at zone, route and route network levels (10 - 20 days) -->
<!-- - Extend the analysis to support prioritisation of other types of intervention beyond new cycleway provision, specifically spatial network analysis using the `igraph` software to identify residential roads on which there is likely to be unsafe through traffic to prioritise 'filtered permeability' interventions (15 - 30 days) -->

<!-- Inception meeting: -->

<!-- Aim is to get the contract moving -->

<!-- Kevin + Rabina + John -->

<!-- Question for Kevin: how involved do you want to be? -->

<!-- Happy to be involved in weekly catch-up -->

<!-- Governance: weekly catch-ups -->

<!-- On the report: it's going to help LAs make big decisions quickly. -->
<!-- Maximum 3-4 pages in plain English -->

<!-- What have we done -->

<!-- How can it help -->

<!-- Limitations -->

<!-- Clear, simple. -->

<!-- Dissemination: we should create a website where we can publish the results - suggestion... popupactive.bike -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, out.width = "100%", eval = FALSE)
```
